title: "variable selection "
output: pdf_document
date: "2025-12-10"
---


``````{r knitr::opts_chunk$set(echo = TRUE)}
```

```{r setup, include=FALSE}
# Set global options and notebook root directory
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Asus.B990/Desktop/Regrassion Project")

```


```{r load-data, echo=TRUE}
data <- read.csv("StudentPerformanceFactors.csv")
if(interactive()) View(data)
```


```{r}
############################################################
# 1. LOAD DATA
############################################################
library(readr)

data <- read_csv("StudentPerformanceFactors.csv",, show_col_types = FALSE)
Sample <- data   # original copy

############################################################
# 2. REMOVE OUTLIERS (your provided list)
############################################################
Sample3 <- Sample[-c(95, 218, 405, 530, 559, 561, 638, 771, 837, 920, 1100, 1108, 
                     1110, 1352, 1526, 1608, 1845, 1864, 2077, 2293, 2422, 2426, 
                     2514, 2543, 2596, 2688, 2905, 2955, 3125, 3142, 3365, 3458, 
                     3580, 3925, 3933, 4193, 4255, 4298, 4356, 4406, 4532, 4584, 
                     4667, 4780, 5126, 5967, 5990, 6348, 6394, 6523), ]
```

############################################################
# 3. CONVERT QUALITATIVE VARIABLES TO FACTORS
############################################################
```{r}

qual_vars <- c("Parental_Involvement", "Access_to_Resources",
               "Extracurricular_Activities", "Motivation_Level",
               "Internet_Access", "Family_Income", "Teacher_Quality",
               "School_Type", "Peer_Influence", "Learning_Disabilities",
               "Parental_Education_Level", "Distance_from_Home", "Gender")

Sample3[qual_vars] <- lapply(Sample3[qual_vars], factor)

Sample3 <- na.omit(Sample3)
```

############################################################
# CHUNK 4: BUILD THE FULL MODEL (OUTLIER REMOVED)
############################################################
# Convert categorical variables into dummy factors for model2
```{r}
dummyParental_I <- factor(Sample3$Parental_Involvement)
dummyAccess_to_Resources <- factor(Sample3$Access_to_Resources)
dummyExtracurricular_Activities <- factor(Sample3$Extracurricular_Activities)
dummyMotivation_Level <- factor(Sample3$Motivation_Level)
dummyInternet_Access <- factor(Sample3$Internet_Access)
dummyFamily_Income <- factor(Sample3$Family_Income)
dummyTeacher_Quality <- factor(Sample3$Teacher_Quality)
dummySchool_Type <- factor(Sample3$School_Type)
dummyPeer_Influence <- factor(Sample3$Peer_Influence)
dummyLearning_Disabilities <- factor(Sample3$Learning_Disabilities)
dummyParental_Education_Level <- factor(Sample3$Parental_Education_Level)
dummyDistance_from_Home <- factor(Sample3$Distance_from_Home)
dummyGender <- factor(Sample3$Gender)

v1 <- Sample3$Hours_Studied
v2 <- Sample3$Attendance
v3 <- Sample3$Sleep_Hours
v4 <- Sample3$Previous_Scores
v5 <- Sample3$Tutoring_Sessions
v6 <- Sample3$Physical_Activity

# Full model after removing outliers
full.Model2 <- lm(Exam_Score ~ dummyParental_I + dummyAccess_to_Resources + 
                    dummyExtracurricular_Activities + dummyMotivation_Level +
                    dummyInternet_Access + dummyFamily_Income + dummyTeacher_Quality +
                    dummySchool_Type + dummyPeer_Influence + dummyLearning_Disabilities +
                    dummyDistance_from_Home + dummyGender +dummyParental_Education_Level+ v1 + v2 + v3 + v4 + v5 + v6,
                  data = Sample3)

cat("=== FULL MODEL AFTER OUTLIER REMOVAL ===\n")
summary(full.Model2)

# Plot model diagnostics
plot(full.Model2)
```

  
  
```{r}
cat("\n=== BACKWARD ELIMINATION MODEL ===\n")
backward.model<-step(full.Model,direction = "backward")



```

############################################################
# 5. MODEL SELECTION
############################################################

```{r}
# 6a. BACKWARD ELIMINATION
backward.model <- step(full.Model2, direction="backward", trace=FALSE)

cat("\n=== BACKWARD ELIMINATION MODEL ===\n")
summary(backward.model)
```

```{r}

# 6b. FORWARD SELECTION
empty.model <- lm(Exam_Score ~ 1, data=Sample3)
forward.model <- step(empty.model,
                      scope = formula(full.Model2),
                      direction = "forward",
                      trace = FALSE)
cat("\n=== FORWARD SELECTION MODEL ===\n")
summary(forward.model)

```

```{r}
# 6c. STEPWISE SELECTION
stepwise.model <- step(full.Model2, direction="both", trace=FALSE)
cat("\n=== STEPWISE MODEL ===\n")
summary(stepwise.model)

```



```{r}

```

############################################################
# 6. COMPARE MODELS (AIC, BIC, Adj-R²)
############################################################
```{r}
############################################################
# CHUNK 7: MODEL COMPARISON
############################################################
model.names <- c("Full", "Backward", "Forward", "Stepwise")

AIC.values <- c(AIC(full.Model2), AIC(backward.model), AIC(forward.model), AIC(stepwise.model))
BIC.values <- c(BIC(full.Model2), BIC(backward.model), BIC(forward.model), BIC(stepwise.model))
AdjR2.values <- c(summary(full.Model2)$adj.r.squared,
                  summary(backward.model)$adj.r.squared,
                  summary(forward.model)$adj.r.squared,
                  summary(stepwise.model)$adj.r.squared)

comparison <- data.frame(Model=model.names,
                         AIC=round(AIC.values,2),
                         BIC=round(BIC.values,2),
                         Adjusted_R2=round(AdjR2.values,4))

cat("\n===== MODEL COMPARISON =====\n")
print(comparison)



```

############################################################
# 7. IDENTIFY BEST MODEL
############################################################
```{r}
best.AIC <- model.names[which.min(AIC.values)]
best.BIC <- model.names[which.min(BIC.values)]
best.AdjR2 <- model.names[which.max(AdjR2.values)]



cat("\n===== FINAL SELECTED MODEL (Stepwise) =====\n")
summary(stepwise.model)
```

















```{r}
library(readr)
Sample <- read_csv("StudentPerformanceFactors.csv")
################################################################
#Sample <- StudentPerformanceFactors
dummyParental_I <- factor(Sample$Parental_Involvement)
dummyAccess_to_Resources<- factor(Sample$Access_to_Resources)
dummyExtracurricular_Activities<- factor(Sample$Extracurricular_Activities)
dummyMotivation_Level<- factor(Sample$Motivation_Level)
dummyInternet_Access<- factor(Sample$Internet_Access)
dummyFamily_Income<- factor(Sample$Family_Income)
dummyTeacher_Quality <- factor(Sample$Teacher_Quality)
dummySchool_Type<- factor(Sample$School_Type)
dummyPeer_Influence<- factor(Sample$Peer_Influence)
dummyLearning_Disabilities<- factor(Sample$Learning_Disabilities)
dummyParental_Education_Level<- factor(Sample$Parental_Education_Level)
dummyDistance_from_Home<- factor(Sample$Distance_from_Home)
dummyGender<- factor(Sample$Gender)
################################################################
v1 <- Sample$Hours_Studied
v2 <- Sample$Attendance
v3 <- Sample$Sleep_Hours
v4 <- Sample$Previous_Scores
v5 <- Sample$Tutoring_Sessions
v6 <- Sample$Physical_Activity
sample2 <- data.frame(v1,v2,v3,v4,v5,v6)
library(corrplot)
cor_matrix <- cor(sample2)
cor_matrix 
corrplot(cor_matrix, method = "circle", addCoef.col = "black",  # adds correlation coefficients as numbers
         tl.col = "black", tl.srt = 45)

######################################################################
full.Model <- lm(Exam_Score ~ dummyParental_I+dummyAccess_to_Resources+dummyExtracurricular_Activities+dummyMotivation_Level+
                 dummyInternet_Access+dummyFamily_Income+dummyTeacher_Quality+dummySchool_Type+dummyPeer_Influence+dummyParental_Education_Level+
                   dummyLearning_Disabilities+dummyDistance_from_Home+dummyGender+v1+v2+v3+v4+v5+v6,data = Sample)
install.packages("car")     
library(car)                
vif(full.Model)
summary(full.Model)
plot(full.Model)


##########################################################
RS <- rstandard(full.Model) 
FM <- fitted(full.Model)          
res <- residuals(full.Model)
```


```{r}
# Standardized residuals
######################################3
# checking influential points
plot(cooks.distance(full.Model),
     pch = 19,
     col = "blue",
     main = "Cook's Distance (fullmodel)",
     ylab = "Cook's Distance",
     xlab = "Index")
abline(h = 4/length(res), col = "red", lty = 2)

cooks <- cooks.distance(full.Model)
max_cooks <- max(cooks)
max_cooks
# indices with |std_res| > 3
out_idx <- which(abs(RS) > 3)
out_idx 

A <- data.frame(Index = out_idx, Residuals = round(RS [out_idx], 6))
A
```


###################################################################
#after removing the outliers

```{r}
Sample3 <- Sample[-c(95, 218, 405, 530, 559, 561, 638, 771, 837, 920, 1100, 1108, 1110, 1352, 1526, 1608, 1845, 1864, 2077, 2293, 2422, 2426, 2514, 2543, 2596, 2688, 2905, 2955, 3125, 3142, 3365, 3458, 3580, 3925, 3933, 4193, 4255, 4298, 4356, 4406, 4532, 4584, 4667, 4780, 5126, 5967, 5990, 6348, 6394, 
                     6523), ]

dummyParental_I <- factor(Sample3$Parental_Involvement)
dummyAccess_to_Resources<- factor(Sample3$Access_to_Resources)
dummyExtracurricular_Activities<- factor(Sample3$Extracurricular_Activities)
dummyMotivation_Level<- factor(Sample3$Motivation_Level)
dummyInternet_Access<- factor(Sample3$Internet_Access)
dummyFamily_Income<- factor(Sample3$Family_Income)
dummyTeacher_Quality <- factor(Sample3$Teacher_Quality)
dummySchool_Type<- factor(Sample3$School_Type)
dummyPeer_Influence<- factor(Sample3$Peer_Influence)
dummyLearning_Disabilities<- factor(Sample3$Learning_Disabilities)
dummyParental_Education_Level<- factor(Sample3$Parental_Education_Level)
dummyDistance_from_Home<- factor(Sample3$Distance_from_Home)
dummyGender<- factor(Sample3$Gender)


Sample3$Parental_Involvement       <- factor(Sample3$Parental_Involvement)
Sample3$Access_to_Resources        <- factor(Sample3$Access_to_Resources)
Sample3$Extracurricular_Activities <- factor(Sample3$Extracurricular_Activities)
Sample3$Motivation_Level           <- factor(Sample3$Motivation_Level)
Sample3$Internet_Access            <- factor(Sample3$Internet_Access)
Sample3$Family_Income              <- factor(Sample3$Family_Income)
Sample3$Teacher_Quality            <- factor(Sample3$Teacher_Quality)
Sample3$School_Type                <- factor(Sample3$School_Type)
Sample3$Peer_Influence             <- factor(Sample3$Peer_Influence)
Sample3$Learning_Disabilities      <- factor(Sample3$Learning_Disabilities)
Sample3$Parental_Education_Level   <- factor(Sample3$Parental_Education_Level)
Sample3$Distance_from_Home         <- factor(Sample3$Distance_from_Home)
Sample3$Gender                     <- factor(Sample3$Gender)




v1 <- Sample3$Hours_Studied
v2 <- Sample3$Attendance
v3 <- Sample3$Sleep_Hours
v4 <- Sample3$Previous_Scores
v5 <- Sample3$Tutoring_Sessions
v6 <- Sample3$Physical_Activity
#####################################################################
#model after removing the outliers seems more stable
#full.Model2 <- lm(Exam_Score ~ dummyParental_I+dummyAccess_to_Resources+dummyExtracurricular_Activities+dummyMotivation_Level+
                   #dummyInternet_Access+dummyFamily_Income+dummyTeacher_Quality+dummySchool_Type+dum#myPeer_Influence+dummyParental_Education_Level+
                   #dummyLearning_Disabilities+dummyDistance_from_Home+dummyGender+v1+v2+v3+v4+v5+v6,data = Sample3)


full.Model2 <- lm(
  Exam_Score ~ Parental_Involvement + Access_to_Resources +
    Extracurricular_Activities + Motivation_Level + Internet_Access +
    Family_Income + Teacher_Quality + School_Type + Peer_Influence +
    Parental_Education_Level + Learning_Disabilities + Distance_from_Home +
    Gender + Hours_Studied + Attendance + Sleep_Hours + Previous_Scores +
    Tutoring_Sessions + Physical_Activity,
  data = Sample3
)

summary(full.Model2)
plot(full.Model2)
vif(full.Model2)

#############################################################
#Checking the assumptions
#Residuals look normal
#It looks like we have some heteroskedasticity but I cant tell because of the categorical varibles is it suppose to look like this?

residuals(full.Model2)
RS=rstandard(full.Model2)
FM=fitted(full.Model2)
hist(RS,xlab="Standard Resduals",main="Histogram of standard Resdual",col=4,freq = FALSE ,breaks = 11,prob=T,ylim = c(0,1))
lines(density(RS),col=1,lwd=3)
####################################################
#assumption 1,2
plot(RS~FM,xlab="Fitted Values",ylab="Standard Resduals",pch=20,cex=3,col=2,main="standard residuals VS fitted values from the estimated regression model")
abline(h=0,lty=10)

#assumption 4
qqPlot(RS, main = "Normal QQ Plot of Residuals")
qqnorm(RS,col=4,cex=2,pch=19)
qqline(RS,lty=2,col=1,cex=4,pch=20)
hist(RS,xlab="Standard Resduals",main="Histogram of standard Resdual",col=4,freq = FALSE ,breaks = 11,prob=T,ylim = c(0,1))
lines(density(RS),col=1,lwd=3)

#Outlier check
plot(RS,
     pch = 19,
     col = "darkgreen",
     main = "Standardized Residuals Index Plot (full model)",
     ylab = "Standardized Residuals",
     xlab = "Index")
abline(h = c(-3, 3), col = "red", lty = 2)
#########################################################################################################
```
```



